---
title: "mixture HBMNL fit (2 chains) — bayesm (rhierMnlRwMixture), saves full posterior"
format:
  html:
    toc: true
    number-sections: true
engine: knitr
params:
    JOB_DIR: NULL
    JOB_ROW: 0          # zero-indexed row of params.csv
    JOB_TESTING: true
---

Fits ONE overspecified (K_MODEL = 5) hierarchical Bayesian mixture multinomial logit
model to one simulated dataset with **bayesm::rhierMnlRwMixture** (Gibbs + random-walk
Metropolis), then saves the **full posterior** in the SAME canonical format the Liesel
jobs (101-103) produce, so the identical `analysis/post_process.py` pipeline (ECR,
marginal comparison on the full + Chebyshev grids, recovery tables, moments) works
unchanged. Shared code is vendored from HierarchicalBayesianMultinomialLogit @ 893e63f.

Engine split: R does pure sampling and writes raw column-major draws; reticulate-Python
rebuilds the TFP `FillScaleTriL` latent from the exported precision and assembles the
canonical `posterior_raw.pkl`. CAPSLOCK variables are harness machinery — don't rename.

## Setup

### Resolve JOB_DIR (works in RStudio, VS Code/Positron, and the SLURM CLI render)

```{r}
#| label: get-default-params
if (!exists("params")) {
  current_file <- rstudioapi::getActiveDocumentContext()$path
  knitr_params <- knitr::knit_params(readr::read_file(current_file))
  params <- list()
  for (param in names(knitr_params)) {
    if (param == "JOB_DIR") {
      current_dir <- fs::path_dir(current_file)
      base_dir <- fs::path(current_dir, "..", "..") |> fs::path_real()
      params[["JOB_DIR"]] <- fs::path_rel(current_dir, start = base_dir)
    } else {
      params[[param]] <- knitr_params[[param]]$value
    }
  }
}
```

```{r}
#| label: get-jobdir
if (is.null(params$JOB_DIR)) {
  current_dir <- knitr::current_input(dir = TRUE) |> fs::path_dir()
  base_dir <- fs::path(current_dir, "..", "..") |> fs::path_real()
  JOB_DIR <- fs::path_rel(current_dir, start = base_dir)
} else {
  JOB_DIR <- params$JOB_DIR
}
```

### Harness machinery (params row, directories, logger)

```{python}
#| label: imports
import json
import logging
import pickle
import shutil
import sys
import tempfile
import time
from datetime import datetime
from pathlib import Path

import numpy as np
import pandas as pd

print("python:", sys.executable)
print("cwd   :", Path.cwd())
```

```{python}
#| label: load-params-row
JOB_DIR: str = r.JOB_DIR
JOB_ROW: int = int(r.params["JOB_ROW"])
JOB_TESTING: bool = bool(r.params["JOB_TESTING"])

JOB_PATH = Path(JOB_DIR)
PARAMS = pd.read_csv(JOB_PATH / "params.csv").iloc[JOB_ROW, :]
print(PARAMS)
```

```{python}
#| label: identifier-and-dirs
now = datetime.now().strftime("%Y-%m-%d-%H%M")
JOB_IDENTIFIER = JOB_PATH.name + "-" + f"{JOB_ROW:04d}" + "-" + now
PARAMS["job"] = JOB_PATH.name
PARAMS["run"] = JOB_IDENTIFIER

DIR_LOG = JOB_PATH / "log"
DIR_OUT = JOB_PATH / "out"
DIR_FIN = JOB_PATH / "finished"
if JOB_TESTING:
    DIR_OUT = JOB_PATH / "out-test"
for d in (DIR_LOG, DIR_OUT, DIR_FIN):
    d.mkdir(exist_ok=True)

FINFILE = DIR_FIN / str(JOB_ROW)
if FINFILE.exists() and not JOB_TESTING:
    raise RuntimeError(
        f"Row {JOB_ROW} of '{JOB_PATH.name}' is already finished. Delete {FINFILE} to re-run."
    )

formatter = logging.Formatter("%(asctime)s - %(levelname)s - %(message)s")
handler = logging.FileHandler(filename=DIR_LOG / f"run-{JOB_ROW:04d}.log")
handler.setFormatter(formatter)
logger = logging.getLogger(str(JOB_PATH.name))
logger.addHandler(handler)
logger.setLevel(logging.INFO)
logger.info(f"Job started: {JOB_IDENTIFIER}")
```

### Extract run settings + resolve shared src/ and data/

The shared `src/` is at the repo root locally (`JOB_PATH/../../src`) and symlinked into
the node-local rundir on the HPC (`JOB_PATH/src`). `data/in/` is two levels above the job
dir either way (see `template.sh.j2`). Resolve both robustly.

```{python}
#| label: run-settings
JOB_PATH_ABS = JOB_PATH.resolve()

SRC_DIR = next((c for c in (JOB_PATH_ABS / "src", JOB_PATH_ABS.parent.parent / "src")
                if (c / "bayesm_sampler.R").exists()), None)
if SRC_DIR is None:
    raise FileNotFoundError("Could not locate src/bayesm_sampler.R next to the job or at the repo root.")
sys.path.insert(0, str(SRC_DIR.parent))           # so `import src.bayesm_convert` resolves

dataset_key = str(PARAMS["dataset_key"])
scenario    = str(PARAMS.get("scenario", "NA"))
k_model     = int(PARAMS["k_model"])
n_chains    = int(PARAMS["n_chains"])
r_total     = int(PARAMS["r_total"])
burn_in     = int(PARAMS["burn_in"])
thin        = int(PARAMS["thin"])
seed        = int(PARAMS["seed"])
a_mu        = float(PARAMS["a_mu"])
a_delta     = float(PARAMS["a_delta"])
dirichlet_a = float(PARAMS["dirichlet_a"])

# Testing shortcut: a short-but-valid chain (still > a few hundred retained draws).
if JOB_TESTING:
    r_total, burn_in, thin = 2000, 200, 2          # -> (2000-200)/2 = 900 retained
    logger.info(f"JOB_TESTING: r_total={r_total}, burn_in={burn_in}, thin={thin}")

DATA_DIR = JOB_PATH_ABS.parent.parent / "data" / "in" / "k5model_mixture"
data_path = str(DATA_DIR / f"{dataset_key}.json")
with open(data_path) as f:
    raw = json.load(f)
P = int(raw["n_params"])
k_true = int(raw["K"])

RUN_KEY = f"{dataset_key}__k{k_model}_bayesm_c{n_chains}"
RAW_DIR = tempfile.mkdtemp(prefix=f"bayesm_{RUN_KEY}_")   # node-local scratch for the .bin bridge
SRC_DIR_STR = str(SRC_DIR)

print(f"dataset={dataset_key}  K_true={k_true}  N={raw['n_units']}  P={P}")
print(f"sampler=bayesm  K_model={k_model}  chains={n_chains}  r_total={r_total}  burn_in={burn_in}  thin={thin}")
print(f"raw bridge dir: {RAW_DIR}")
```

## Model code — bayesm sampling (R)

Source the vendored sampler and fit. Settings come from the Python block via `py$`;
the result is written to `RAW_DIR` as `.bin` draws + `dims.json` + `meta_r.json`.

```{r}
#| label: bayesm-fit
library(reticulate)   # attaches the `py` object so R can read the Python session's vars
source(file.path(py$SRC_DIR_STR, "bayesm_sampler.R"))
run_bayesm_fit(
  data_path   = py$data_path,
  out_raw_dir = py$RAW_DIR,
  k_model     = py$k_model,
  chains      = py$n_chains,
  r_total     = py$r_total,
  burn_in     = py$burn_in,
  thin        = py$thin,
  seed        = py$seed,
  a_mu        = py$a_mu,
  a_delta     = py$a_delta,
  dirichlet_a = py$dirichlet_a,
  scenario    = py$scenario
)
```

## Save the full posterior + meta (canonical format, identical to 004-007)

```{python}
#| label: convert-and-save
from src.bayesm_convert import read_raw_draws

post_np, meta_r, dims = read_raw_draws(RAW_DIR)
runtime_s = float(meta_r.get("total_duration_s") or 0.0)

# Tier-2 subset: keep the FULL posterior only for the first few replicate seeds (needed for
# the cross-sampler marginal-density comparison). Every run still writes the tidy per-run
# summaries below; the rest are summarised on-node and the posterior discarded, so the study
# scales to many seeds without shipping GBs. See src/summaries.py.
POSTERIOR_SUBSET_MAX = 5
data_seed = int(PARAMS.get("data_seed", -1))
SAVE_POSTERIOR = JOB_TESTING or data_seed <= POSTERIOR_SUBSET_MAX
if SAVE_POSTERIOR:
    DIR_POST = DIR_OUT / "posterior_raw"
    DIR_POST.mkdir(parents=True, exist_ok=True)
    with open(DIR_POST / f"{RUN_KEY}.pkl", "wb") as f:
        pickle.dump(post_np, f)
    print("saved posterior_raw:", {k: tuple(v.shape) for k, v in post_np.items()})
else:
    print(f"posterior_raw NOT saved (data_seed={data_seed} > {POSTERIOR_SUBSET_MAX}); on-node summaries only")

shutil.rmtree(RAW_DIR, ignore_errors=True)          # drop the large .bin bridge files
```

```{python}
#| label: save-meta
# Cheap label-invariant convergence (no beta_i) for at-a-glance monitoring — same as 004-007.
try:
    from src import analysis
    conv = analysis.invariant_convergence_summary(post_np, include_cov=True)
    rhat_max = float(np.nanmax(conv["rhat"].values))
    ess_min = float(np.nanmin(conv["ess"].values))
except Exception as e:
    logger.warning(f"invariant convergence failed: {e}")
    rhat_max, ess_min = None, None

# COND_KEYS must be byte-identical to the Liesel meta so post_process discovers bayesm runs
# (dataset_key, scenario, k_true, data_seed, k_model, sampler, n_chains). 'sampler' = bayesm.
meta = {
    "job": str(PARAMS["job"]), "run": str(PARAMS["run"]), "job_row": int(JOB_ROW),
    "run_key": RUN_KEY,
    "dataset_key": dataset_key, "scenario": scenario,
    "k_true": k_true, "data_seed": int(PARAMS.get("data_seed", -1)),
    "k_model": k_model, "sampler": "bayesm", "n_chains": n_chains,
    "warmup": burn_in, "posterior": int(dims["n_samples"]), "seed": seed,
    "a_mu": a_mu, "a_delta": a_delta, "dirichlet_a": dirichlet_a,
    "num_integration_steps": None,
    "n_units": int(raw["n_units"]), "n_params": P, "n_demos": int(raw.get("n_demos", 0)),
    "runtime_s": round(runtime_s, 2),
    "n_sampling_errors": 0,            # bayesm has no divergence/error log (Gibbs+RW-Metropolis)
    "invariant_rhat_max": rhat_max, "invariant_ess_min": ess_min,
    "posterior_shapes": {k: list(v.shape) for k, v in post_np.items()},
    # bayesm-specific provenance
    "r_total": r_total, "thin": thin, "burn_in": burn_in,
    "seeds": meta_r.get("seeds"), "chain_durations_s": meta_r.get("durations_s"),
    "loglike_mean": meta_r.get("loglike_mean"),
    "bayesm_version": meta_r.get("bayesm_version"), "r_version": meta_r.get("r_version"),
    "finished_at": datetime.now().isoformat(timespec="seconds"),
}
DIR_META = DIR_OUT / "meta"
DIR_META.mkdir(parents=True, exist_ok=True)
with open(DIR_META / f"{RUN_KEY}.json", "w") as f:
    json.dump(meta, f, indent=2)
print("meta:", json.dumps({k: meta[k] for k in
      ("run_key", "runtime_s", "invariant_rhat_max", "invariant_ess_min", "bayesm_version")}))
```

## Save per-run summary tables (Tier 1)

Every tidy table that can be computed from THIS run's posterior + ground truth is written
here as `out/<table>/<run_key>.csv`: recovery of `Delta` / `beta` / `mu` / `Sigma` /
weights, the ECR report, convergence and mixture moments. The same `src/summaries.py`
builds them on-node here AND post-hoc in `analysis/post_process.py`, so the per-run CSVs
and any gathered table agree exactly. This is what lets the study scale to many seeds: the
headline metrics ride along as KB-sized CSVs while only the Tier-2 subset keeps the full
posterior. If summarising fails, the posterior is saved as a fallback so an expensive run
is never lost.

```{python}
#| label: save-summaries
from src import summaries as smry

# bayesm (Gibbs + RW-Metropolis) has no per-transition diagnostics -> no diagnostics table.
try:
    summary_tables, _ = smry.per_run_tables(post_np, meta, raw, diag=None)
    for _name, _rows in summary_tables.items():
        if not _rows:
            continue
        _dir = DIR_OUT / _name
        _dir.mkdir(parents=True, exist_ok=True)
        pd.DataFrame(_rows).to_csv(_dir / f"{RUN_KEY}.csv", index=False)
    logger.info("per-run summaries: " + ", ".join(f"{k}={len(v)}" for k, v in summary_tables.items()))
    print("saved summaries:", {k: len(v) for k, v in summary_tables.items()})
except Exception as e:  # never lose a run to a summary bug: keep the posterior to re-derive
    logger.error(f"on-node summaries failed: {e}")
    if not SAVE_POSTERIOR:
        DIR_POST = DIR_OUT / "posterior_raw"
        DIR_POST.mkdir(parents=True, exist_ok=True)
        with open(DIR_POST / f"{RUN_KEY}.pkl", "wb") as f:
            pickle.dump(post_np, f)
        logger.error("saved posterior_raw as fallback for post-hoc re-derivation")
    print(f"WARNING: on-node summaries failed ({e}); see log")
```

## Mark job as finished

```{python}
#| label: mark-finished
if not JOB_TESTING:
    FINFILE.touch()
    logger.info("Marked finished.")
```
